---
title: "SpectralPlusDiffusion"
subtitle: "Course Project for Math322"
author: "Ruiqi Li ruiqi.li@yale.edu"
date: "12/14/2022"
output: pdf_document
---
# Package
```{r}
library(dirmult)
library(splatter)
library(scater)
library(cluster)
library(Seurat)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(aricode)
library(foreach)
library(doParallel)
```

# Function
```{r}
source("./SPD_func.R")
```

# Simulation Data
```{r}
params <- newSplatParams()
nGroups = 9
set.seed(100)
group.prob = as.numeric(rdirichlet(n=1, alpha = rep(1,nGroups)))
params <- setParams(params, update = list(
  mean.rate = 0.3,
  mean.shape = 0.6,
  lib.loc = 11,
  lib.scale = 0.2,
  lib.norm = FALSE,
  out.prob = 0.05,
  out.facLoc = 4,
  out.facScale = 0.5,
  group.prob = group.prob,
  nGenes = 13714, 
  batchCells = 2638))

seurat_obj <- splatSimulate(params, method = "group")
seurat_obj <- logNormCounts(seurat_obj)
seurat_obj <- as.Seurat(seurat_obj, counts = "counts")
seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_obj <- ScaleData(seurat_obj, features = rownames(seurat_obj))
seurat_obj <- RunPCA(seurat_obj, features = rownames(seurat_obj))
seurat_obj <- RunUMAP(seurat_obj, dims = 1:15)
seurat_obj[["external_label"]] = seurat_obj$Group
```

# Real data (pbmc3k)
```{r}
# SeuratData::InstallData("pbmc3k")
data("pbmc3k.final")
seurat_obj = pbmc3k.final
seurat_obj[["external_label"]] = seurat_obj$seurat_annotations
```

# Cluster with external label
```{r}
p = DimPlot(seurat_obj, group.by = "external_label", label = TRUE)
p
```

# Obtain Gaussian Kernel
```{r}
delta = 0.1
kNN = 10
nbEigenValues_spec = 50 # num of eigvals used to do spectral clustering
nbEigenValues_diffuse = 100 # largest num of eigvals for diffusiondist
# eigval_thred = 0.99

dat = as.matrix(seurat_obj@reductions$pca@cell.embeddings)
external_label = seurat_obj$external_label
sim_mat = Build_Gaussian_kernel(data = dat, kNN = kNN)
```

# Clustering by different metric
## Spectral Clustering
```{r}
spectral_cluster = Spectral_clustering(sim_mat, k = 9)
Idents(seurat_obj) = spectral_cluster
p1 = DimPlot(seurat_obj, label = TRUE) + ggtitle(sprintf("Spectral Clustering, ARI = %.2f",round(ARI(external_label, spectral_cluster),2))) + NoLegend()
p1
```

## Hierarchical Clustering & K-medoids using Diffusion distance
### Determine time-scale
```{r}
ari_h = NULL
cl_h = NULL
ari_k = NULL
cl_k = NULL
time_scale_ls = c(seq(1,10),seq(20,50,10))
for(t in time_scale_ls){
  cat(t,"\n")
  diffuse_dist = Calculate_diffusiondist(sim_mat = sim_mat,
                                       nbEigenValues = nbEigenValues_diffuse,
                                       time_scale = t, delta = 0.1)
  htree_res = Hierarchical_clustering(diffusion_dist = diffuse_dist,
                                    k = 9)
  cl_h = c(cl_h,list(htree_res$partition))
  ari_h = c(ari_h,ARI(external_label,htree_res$partition))
  
  kmedoid_res = kmedoid_clustering(diffusion_dist = diffuse_dist,
                                    k = 9)
  cl_k = c(cl_k,list(kmedoid_res))
  ari_k = c(ari_k,ARI(external_label,kmedoid_res))
}
```
Plot ARI curve
```{r}
df = data.frame("time_scale" = time_scale_ls,"ARI_hclust" = ari_h, "ARI_kmedoid" = ari_k)
df$time_scale = factor(df$time_scale,levels = unique(df$time_scale))
df = reshape2::melt(df, id="time_scale") 
colnames(df) = c("time_scale","method","ARI")

ggplot(data=df, 
       aes(x=time_scale, y=ARI,group = method,color = method)) + 
  geom_line()+ geom_point()
```

### Hierachical Clustering
```{r}
t = min(which(ari_h == max(ari_h)))
Idents(seurat_obj) = cl_h[[t]]
p2 = DimPlot(seurat_obj, label = TRUE) + ggtitle(sprintf("Hierachical Clustering (t = %d), ARI = %.2f",t,round(ARI(external_label, cl_h[[t]]),2))) + NoLegend()
p2
```

### K-medoids Clustering
```{r}
t = min(which(ari_k == max(ari_k)))
Idents(seurat_obj) = cl_k[[t]]
p2 = DimPlot(seurat_obj, label = TRUE) + ggtitle(sprintf("Hierachical Clustering (t = %d), ARI = %.2f",t,round(ARI(external_label, cl_k[[t]]),2))) + NoLegend()
p2
```


# Spectral + K-medoids Clustering
## Spectral Clustering
### Define the number of clusters by eigenval decaying
```{r}
Vals = ncut(W = sim_mat, nbEigenValues = 50)[[2]]
plot(Vals)
abline(h = 0.98, col = "red")
```

Here we set cutoff at eigenVal = 0.98, i.e K = 4.

### Performing spectral clustering to separate disconnected components apart
```{r}
spectral_cluster = Spectral_clustering(sim_mat, k = 4)
Idents(seurat_obj) = spectral_cluster
DimPlot(seurat_obj, label = TRUE) + ggtitle(paste0("Spectral Clustering, ARI = ",round(ARI(external_label, spectral_cluster),2))) + NoLegend()
```

## K-medoid on each subgroup
### Determine t by twin sample cross-validation
```{r}
time_scale_ls = c(seq(1,10),seq(20,50,10))
cl <- makeCluster(10)
registerDoParallel(cl)
res <- foreach(i = time_scale_ls, .packages = c("kmed", "aricode"), .combine = rbind) %dopar% {
  res = unlist(lapply(names(table(spectral_cluster)), function(part){
    subsample = which(spectral_cluster == part)
    sim_mat_sub = sim_mat[subsample,subsample]
    diffuse_dist = Calculate_diffusiondist(sim_mat = sim_mat_sub,
                                       nbEigenValues = nbEigenValues_diffuse,
                                       time_scale = i, delta = 0.1)
    RI = unlist(lapply(1:10,function(seed){
      Twin_sample_metric(diffuse_dist, seed = seed)
    }) )
    mean(RI)
  }) )
  
  return(res)
}
stopCluster(cl)
registerDoSEQ()
```

```{r}
# Plot average Rand Idex curve
df = data.frame(res)
df$'time_scale' = time_scale_ls
colnames(df)[1:4] = names(table(spectral_cluster))
df = reshape2::melt(df,id = "time_scale")
colnames(df) = c("time_scale","cluster","RI")

df$time_scale = factor(df$time_scale,levels = unique(df$time_scale))
ggplot(data=df, 
       aes(x=time_scale, y=RI,group = cluster,color = cluster)) + 
  geom_line()+ geom_point()
```
Here we choose t = 10.

### Calculate the diffusion distance for each group
```{r}
t = 10
dist_ls = lapply(names(table(spectral_cluster)), function(part){
  subsample = which(spectral_cluster == part)
  sim_mat_sub = sim_mat[subsample,subsample]
  diffuse_dist = Calculate_diffusiondist(sim_mat = sim_mat_sub,
                                   nbEigenValues = nbEigenValues_diffuse,
                                   time_scale = t, delta = 0.1)
  diffuse_dist
})

names(dist_ls) = names(table(spectral_cluster))
```

### k-medoid clustering for each group (given k)
```{r}
singleton = 10
k = c(1,1,3,4)
final_cluster = spectral_cluster

for(i in 1:length(dist_ls)){
  subsample = which(spectral_cluster == names(dist_ls)[i])
  
  # discard group w/ singleton
  group_num = 1; k_ = k[i]
  partition = NULL
  while(group_num!=k[i]){
    # partition = Hierarchical_clustering(dist_ls[[i]], k = k_)[[1]]
    partition = kmedoid_clustering(dist_ls[[i]], k = k_)
    group_num = sum(table(partition) > singleton)
    k_ = k_ + 1
  }
  
  final_cluster[subsample] = paste0(spectral_cluster[subsample],"-",partition)
}
```

### Visualization
```{r}
final_cluster = as.character(final_cluster)
# Remove singleton 
singleton_group = names(table(final_cluster)[table(final_cluster) < singleton])
final_cluster[final_cluster %in% singleton_group] = "Outlier"
Idents(seurat_obj) = final_cluster
p3 = DimPlot(seurat_obj, label = TRUE) + ggtitle(sprintf("K-medoid Clustering (t = %d), ARI = %.2f",t,round(ARI(external_label, Idents(seurat_obj)),2)))
p3
```


